---
title: "CMM535 Coursework"
author: "Abimbola"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Importing neccesary libaries

```{r}
library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)
library(caTools)
```

Reading the escapes data and displaying the summary statistics.

```{r}
escapes_df <- read.csv('~/Documents/escapes.csv')
escapes_df
summary(escapes_df)
```
#Task 1

Selecting the variables Escape species, Average Weight, Age, Site.Name as areas of special interest.

```{r}
escapes_df <- escapes_df %>%
                select(Escaped.Species, Age, Average.Weight, Site.Name)

```

Data Cleaning I - Average.Weight

```{r}
# pre-processing the weights in grams (converting all to kilograms)
gram_weights <- escapes_df %>%
                  filter(!grepl('kg', Average.Weight)) %>%
                  filter(grepl('g|grams', Average.Weight)) %>%
                  select(Average.Weight) %>%
                  mutate_all(~gsub('g|grams', '', .))

# averaging the hyphened weights
gram_weights_hyphen <- gram_weights %>%
                          filter(grepl('-', Average.Weight))

low_upper_weights <- gram_weights_hyphen %>%
                          separate(Average.Weight, c('low_lim', 'upper_lim'), sep = '-')

low_upper_weights <- low_upper_weights%>%
                          transform(low_lim = as.numeric(low_lim)) %>%
                          transform(upper_lim = as.numeric(upper_lim))

low_upper_weights$weight <- (low_upper_weights$low_lim + low_upper_weights$upper_lim)/2


gram_weights[grepl('-', gram_weights$Average.Weight),
             'Average.Weight'] <- low_upper_weights$weight

gram_weights$Average.Weight <- as.numeric(gram_weights$Average.Weight)/1000

escapes_df[!grepl('kg', escapes_df$Average.Weight) &
           grepl('g', escapes_df$Average.Weight), 
           'Average.Weight'] <- gram_weights$Average.Weight


# pre-processing the weights in kg (units mentioned)
kgram_weights <- escapes_df %>%
                  filter(grepl('kg|kilos', Average.Weight)) %>%
                  select(Average.Weight) %>%
                  mutate_all(~gsub('kg|kilos', '', .))

# averaging the hyphened weights
kgram_weights_hyphen <- kgram_weights %>%
                          filter(grepl('-', Average.Weight))

low_upper_weights <- kgram_weights_hyphen %>%
                          separate(Average.Weight, c('low_lim', 'upper_lim'), sep = '-')

low_upper_weights <- low_upper_weights%>%
                          transform(low_lim = as.numeric(low_lim)) %>%
                          transform(upper_lim = as.numeric(upper_lim))

low_upper_weights$weight <- (low_upper_weights$low_lim + low_upper_weights$upper_lim)/2

kgram_weights[grepl('-', kgram_weights$Average.Weight),
             'Average.Weight'] <- low_upper_weights$weight

escapes_df[grepl('kg|kilos', escapes_df$Average.Weight),
           'Average.Weight'] <- kgram_weights$Average.Weight

# removing all missing/bad values (NA)
escapes_df <- escapes_df[!is.na(escapes_df$Average.Weight),]

# further pre-processing into numbers and converting the weight in lb to kilograms
escapes_df[escapes_df$Average.Weight == '~1.0 ',] <- 1.0
escapes_df[escapes_df$Average.Weight == '1 lb',] <- 0.45

# converting all the weights to numeric and removing bad values
escapes_df$Average.Weight <- as.numeric(escapes_df$Average.Weight)
escapes_df <- escapes_df[!is.na(escapes_df$Average.Weight),]
```
```{r}
escapes_df
summary(escapes_df)
```


Data cleaning II - Age

```{r}
# pre-processing the ages in months (mentioned)
months_age <- escapes_df %>%
                  filter(grepl('month|months|months+|mo|mths', Age)) %>%
                  select(Age) %>%
                  mutate_all(~gsub('month|months|months+|mo|mths', '', .))

# averaging the hyphened ages
months_age_hyphen <- months_age %>%
                      filter(grepl('-', Age))

low_upper_months <- months_age_hyphen %>%
                          separate(Age, c('low_lim', 'upper_lim'), sep = '-')

low_upper_months <- low_upper_months%>%
                          transform(low_lim = as.numeric(low_lim)) %>%
                          transform(upper_lim = as.numeric(upper_lim))

low_upper_months$age <- (low_upper_months$low_lim + low_upper_months$upper_lim)/2

months_age[grepl('-', months_age$Age),
             'Age'] <- low_upper_months$age

escapes_df[grepl('month|months|months+|mo|mths', escapes_df$Age), 
           'Age'] <- months_age$Age

# pre-processing the ages in years (converting into months)
yrs_age <- escapes_df %>%
            filter(grepl('yrs', Age)) %>%
            select(Age) %>%
            mutate_all(~gsub('yrs', '', .))

yrs_age$Age <- as.numeric(yrs_age$Age)

escapes_df[grepl('yrs', escapes_df$Age), 'Age'] <- yrs_age$Age*12

escapes_df$Age <- as.numeric(escapes_df$Age)

# dropping the bad numerical value, 1999
escapes_df = escapes_df[escapes_df$Age != 1999,]

# removing all missing/bad values (NA) 
escapes_df <- escapes_df[!is.na(escapes_df$Age),]
```
```{r}
escapes_df
summary(escapes_df)
```



Reading the analysis data and displaying the summary statistics.

```{r}
analysis_df <- read.csv('~/Documents/analysis.csv')
analysis_df
summary(analysis_df)
```

Removing the the variables 'year' and 'month' from the data.

```{r}
analysis_df <- analysis_df %>%
                  select(-year, -month) %>%
                  mutate(Site.Name = tolower(Site.Name))
```

#Task 2

Integrating (merging using INNER JOIN on Site.Name) the 2 dataframes, escapes_df (from escapes.csv) and analysis_df (analysis.csv) into escapesPlus dataframe.

```{r}
escapesPlus <- merge(escapes_df, analysis_df, by = 'Site.Name')

escapesPlus

write.csv(escapesPlus, 'escapesPlus.csv', row.names = FALSE)
```
#Task 3

Exporatory Data Analysis on the dataframe, escapesPlus:

1. Mean Age of fishes belonging to every species escaped. (Bar-Plot)
2. Mean Age of fishes belonging to each Average Weight of fishes escaped. (Line-Plot)

```{r}
escapesPlus <- escapesPlus%>%
                          select(-Site.Name)

escapesPlus %>%
  group_by(Escaped.Species) %>%
  summarise(Mean_Age = mean(Age)) %>%
  ggplot(aes(x = Escaped.Species, y = Mean_Age)) +
  geom_bar(stat = 'identity', fill = 'red')

escapesPlus %>%
  group_by(Average.Weight) %>%
  summarise(Mean_Age = mean(Age)) %>%
  ggplot(aes(x = Average.Weight, y = Mean_Age)) +
  geom_line(color = 'red')
```
#Task 4

Developing a Linear Regression Model to predict the Age of the fish escaped given it's species, Average Weight and analysis variables: c2, c3, c4, c5, c6 and c7

```{r}
# 80-20 Train Test Split into Training and Test Sets respectively
set.seed(123)
train_combination <- sample.split(escapesPlus, SplitRatio = 0.8)
train <- escapesPlus[train_combination, ]
test <- escapesPlus[!train_combination, ]

# Training the Linear Regression Model
age_prediction_lm <- lm(Age ~ ., data = train)
summary(age_prediction_lm)

# getting the predictions of the Test Set and Model Performance on Test Set (in terms of R-squared Score)
test_predictions = predict(age_prediction_lm, newdata = test)
print(paste0('Test R-squared Score: ', cor(test$Age, test_predictions)**2))
```

